Commutative character sheaves and geometric types for supercuspidal representations

### 2021-01-06 Geometric types.tex
Final referee changes

As we recalled in Section~\ref{ssec:review}, a Yu type datum consists of compact groups $\oK^i$, a representation $\orho^0$ of $\oK^0$ and quasicharacters $\varphi^i$ of $\oK^i$, for $i=0,\ldots, d$; see \ref{Y0}, \ref{Y1} and \ref{Y2}.
In Section~\ref{ssec:review} we also saw how a Yu type datum determines smooth group schemes $\underline{G}^i$, a representation $\varrho_0\red$ of the reductive quotient $(\underline{G}^0)\red_{k}$ of the special fibre $\underline{G}^0_{k}$ of $\underline{G}^0$, and quasicharacters $\varphi_i$ of the group of $k$-rational points on the level-$r_i$ Greenberg transform $G_i$ of $\underline{G}^i$; see \ref{M0}, \ref{M1} and \ref{M2}.
%
Theorem~\ref{thm:geotypes} places the following conditions on the Yu type datum:
\begin{enumerate}
%\labitem{H0}{H0} the geometric component group of $(\underline{G}^0)\red_{k}$ is cyclic;
\labitem{H1}{H1} for each $i=0,\ldots, d$, either
\begin{enumerate}
\labitem{H1(a)}{H1a} $G_i$ is reductive, or % with cyclic geometric component group, or
\labitem{H1(b)}{H1b} the quasicharacter $\varphi_i$ of $G_i$ is trivial on  $G_{i,\operatorname{der}}(k)$.
\end{enumerate}
\end{enumerate}
%Hypothesis~\ref{H0} will allow us to use Lusztig's character sheaves to geometrize the character of the representation of $\orho^0$.
Hypothesis~\ref{H1a} may be used to geometrize each quasicharacter $\varphi_i$ using $\mathsf{K}_{\bar\QQ}\mathsf{D}^b_{c,G_i}(G_i;\EE)$.
Alternatively, Hypothesis~\ref{H1b} allows us to use Proposition~\ref{prop:quasicharacters} to geometrize each quasicharacter $\varphi_i$ using local systems on $G_i$, or more precisely, using commutative character sheaves on $G_i$.
Using Sections~\ref{ssec:equivariant} and \ref{ssec:LCS} we see that, in both cases, the resulting geometrization may be interpreted as an element in $\mathsf{K}_{\bar\QQ}\mathsf{D}^b_{c,G_i}(G_i;\EE)$.
Note that when Hypothesis~\ref{H1b} applies, it provides a considerably simpler geometrization than when Hypothesis~\ref{H1a} applies.

We give examples illustrating Hypothesis \ref{H1b}: first,
an example of a quasicharacter that does not vanish on $G_{i, \der}(k)$,
and then a class of Yu data where it is always satisfied.
Let $G = \SL_2$ over $\QQ_2$.
Then the hyperspecial maximal compact $\oK^0$ reduces to $\SL_2(\ZZ/2^{d+1}\ZZ)$ at depth $d$.
The derived subgroup of $\SL_2(\ZZ/2\ZZ)$ has index $2$, while it has index $4$ in positive depth.
There is therefore a character $\varphi^0$ of $K$ of order $4$ and depth $1$
factoring through the quotient of $\SL_2(\ZZ/4\ZZ)$ by its derived subgroup.
Taking the trivial representation for $\orho^0$ gives a datum not satisfying \ref{H1b}.

In contrast, there is a broad class of representations that automatically satisfy \ref{H1b}.
Recall that a representation is \emph{toral} \citelist{\cite{adler:98a} \cite{kaletha:19a}*{\S 6}}
if it arises from a Yu datum of the form $(S \subset G, \orho^0=1, (\varphi^{d-1}, 1)$ where
$S$ is an elliptic maximal torus and $\varphi^{d-1}$ is $G$-generic character of $S(F)$ of positive depth.
Since $S$ is abelian, it has trivial derived subgroup and thus \ref{H1b} is satisfied.
Toral representations include epipelagic representations \cite{reeder-yu:14a} and
the representations considered by Reeder in \cite{reeder:08a}.

@@

\bib{bushnell-kutzko:98a}{article}{
      author={Bushnell, Colin},
      author={Kutzko, Phil},
       title={{Smooth representations of reductive $p$-adic groups: structure
  theory via types}},
        date={1998},
     journal={Proc. London Math. Soc.},
      volume={77},
      number={3},
       pages={{582\ndash 634}},
}

@@

\bib{demazure:SGA3-VIB}{inproceedings}{
      author={Bertin, J.E.},
       title={{Expos\'e VI$_\mathrm{B}$: G\'en\'eralit\'es sur les pr\'esch\'emas en groupes}},
        date={1970},
   booktitle={S\'eminaire de G\'eom\'etrie Alg\'ebrique du Bois Marie, Sch\'emas en groupes I (SGA3) {Proprietes generales des schemas en
  groupes}},
      editor={Demazure, Michel},
      editor={Grothendieck, Alexander},
      series={Lecture Notes in Math},
      volume={151},
   publisher={Springer-Verlag},
     address={Berlin},
}

@@

\bib{lusztig:disconnected4}{article}{
   author={Lusztig, G.},
   title={Character sheaves on disconnected groups. IV},
   journal={Represent. Theory},
   volume={8},
   date={2004},
   pages={145--178},
%   review={\MR{2048590}},
%   doi={10.1090/S1088-4165-04-00240-7},
}

>>>>

As we recalled in Section~\ref{ssec:review}, a Yu type datum consists of compact groups $\oK^i$, a representation $\orho^0$ of $\oK^0$ and quasicharacters $\varphi^i$ of $\oK^i$, for $i=0,\ldots, d$; see \ref{Y0}, \ref{Y1} and \ref{Y2}.
In Section~\ref{ssec:review} we also saw how a Yu type datum determines smooth group schemes $\underline{G}^i$, a representation $\varrho_0\red$ of the reductive quotient $(\underline{G}^0)\red_{k}$ of the special fibre $\underline{G}^0_{k}$ of $\underline{G}^0$, and quasicharacters $\varphi_i$ of the group of $k$-rational points on the level-$r_i$ Greenberg transform $G_i$ of $\underline{G}^i$; see \ref{M0}, \ref{M1} and \ref{M2}.
%
Theorem~\ref{thm:geotypes} places the following conditions on the Yu type datum:
\begin{enumerate}
%\labitem{H0}{H0} the geometric component group of $(\underline{G}^0)\red_{k}$ is cyclic;
\labitem{H1}{H1} for each $i=0,\ldots, d$, either
\begin{enumerate}
\labitem{H1(a)}{H1a} $G_i$ is reductive, or % with cyclic geometric component group, or
\labitem{H1(b)}{H1b} the quasicharacter $\varphi_i$ of $G_i$ is trivial on  $G_{i,\operatorname{der}}(k)$.
\end{enumerate}
\end{enumerate}
%Hypothesis~\ref{H0} will allow us to use Lusztig's character sheaves to geometrize the character of the representation of $\orho^0$.
Hypothesis~\ref{H1a} may be used to geometrize each quasicharacter $\varphi_i$ inside the Grothendieck group $\mathsf{K}_{\bar\QQ}\mathsf{D}^b_{c,G_i}(G_i;\EE)$.
Alternatively, Hypothesis~\ref{H1b} allows us to use Proposition~\ref{prop:quasicharacters} to geometrize each quasicharacter $\varphi_i$ using local systems on $G_i$, or more precisely, using commutative character sheaves on $G_i$.
Using Sections~\ref{ssec:equivariant} and \ref{ssec:LCS} we see that, in both cases, the resulting geometrization may be interpreted as an element in $\mathsf{K}_{\bar\QQ}\mathsf{D}^b_{c,G_i}(G_i;\EE)$.
Note that when Hypothesis~\ref{H1b} applies, it provides a considerably simpler geometrization than when Hypothesis~\ref{H1a} applies.

We give examples illustrating Hypothesis \ref{H1b}: first,
an example of a quasicharacter that does not vanish on $G_{i, \der}(k)$,
and then a class of Yu data where it is always satisfied.
Let $G = \SL_2$ over $\QQ_2$.
Then the hyperspecial maximal compact $\oK^0$ reduces to $\SL_2(\ZZ/2^{d+1}\ZZ)$ at depth $d$.
The derived subgroup of $\SL_2(\ZZ/2\ZZ)$ has index $2$, while it has index $4$ in positive depth.
There is therefore a character $\varphi^0$ of $K$ of order $4$ and depth $1$
factoring through the quotient of $\SL_2(\ZZ/4\ZZ)$ by its derived subgroup.
Taking the trivial representation for $\orho^0$ gives a datum not satisfying \ref{H1b}.
Note that this datum does not satisfy \ref{H1a} either since the depth 1 Greenberg transform
of the hyperspecial $\oK^0$ is not reductive.

In contrast, there is a broad class of representations that automatically satisfy \ref{H1b}.
Recall that a representation is \emph{toral} \citelist{\cite{adler:98a} \cite{kaletha:19a}*{\S 6}}
if it arises from a Yu datum of the form $(S \subset G, \orho^0=1, (\varphi^{d-1}, 1))$ where
$S$ is an elliptic maximal torus and $\varphi^{d-1}$ is $G$-generic character of $S(F)$ of positive depth.
Since $S$ is abelian, it has trivial derived subgroup and thus \ref{H1b} is satisfied.
Toral representations include epipelagic representations \cite{reeder-yu:14a} and
the representations considered by Reeder in \cite{reeder:08a}.

@@

\bib{bushnell-kutzko:98a}{article}{
      author={Bushnell, Colin},
      author={Kutzko, Philip},
       title={{Smooth representations of reductive $p$-adic groups: structure
  theory via types}},
        date={1998},
     journal={Proc. London Math. Soc.},
      volume={77},
      number={3},
       pages={{582\ndash 634}},
}

@@

\bib{demazure:SGA3-VIB}{inproceedings}{
      author={Bertin, Jean-\'Etienne},
       title={{Expos\'e VI$_\mathrm{B}$: G\'en\'eralit\'es sur les pr\'esch\'emas en groupes}},
        date={1970},
   booktitle={S\'eminaire de G\'eom\'etrie Alg\'ebrique du Bois Marie, Sch\'emas en groupes I (SGA3) {Proprietes generales des schemas en
  groupes}},
      editor={Demazure, Michel},
      editor={Grothendieck, Alexander},
      series={Lecture Notes in Math},
      volume={151},
   publisher={Springer-Verlag},
     address={Berlin},
}

@@

\bib{lusztig:disconnected4}{article}{
   author={Lusztig, George},
   title={Character sheaves on disconnected groups. IV},
   journal={Represent. Theory},
   volume={8},
   date={2004},
   pages={145--178},
%   review={\MR{2048590}},
%   doi={10.1090/S1088-4165-04-00240-7},
}
